---
title: "Mini Data-Analysis Deliverable 1"
output: github_document
---

# Welcome to your (maybe) first-ever data analysis project!

And hopefully the first of many. Let's get started:

1.  Install the [`datateachr`](https://github.com/UBC-MDS/datateachr) package by typing the following into your **R terminal**:

<!-- -->

    install.packages("devtools")
    devtools::install_github("UBC-MDS/datateachr")

2.  Load the packages below.

```{r}
library(datateachr)
library(tidyverse)
```

3.  Make a repository in the <https://github.com/stat545ubc-2022> Organization. You will be working with this repository for the entire data analysis project. You can either make it public, or make it private and add the TA's and Lucy as collaborators. A link to help you create a private repository is available on the #collaborative-project Slack channel. 

# Instructions

## For Both Milestones

-   Each milestone is worth 45 points. The number of points allocated to each task will be annotated within each deliverable. Tasks that are more challenging will often be allocated more points.

-   10 points will be allocated to the reproducibility, cleanliness, and coherence of the overall analysis. While the two milestones will be submitted as independent deliverables, the analysis itself is a continuum - think of it as two chapters to a story. Each chapter, or in this case, portion of your analysis, should be easily followed through by someone unfamiliar with the content. [Here](https://swcarpentry.github.io/r-novice-inflammation/06-best-practices-R/) is a good resource for what constitutes "good code". Learning good coding practices early in your career will save you hassle later on!

## For Milestone 1

**To complete this milestone**, edit [this very `.Rmd` file](https://raw.githubusercontent.com/UBC-STAT/stat545.stat.ubc.ca/master/content/mini-project/mini-project-1.Rmd) directly. Fill in the sections that are tagged with `<!--- start your work below --->`.

**To submit this milestone**, make sure to knit this `.Rmd` file to an `.md` file by changing the YAML output settings from `output: html_document` to `output: github_document`. Commit and push all of your work to the mini-analysis GitHub repository you made earlier, and tag a release on GitHub. Then, submit a link to your tagged release on canvas.

**Points**: This milestone is worth 45 points: 43 for your analysis, 1 point for having your Milestone 1 document knit error-free, and 1 point for tagging your release on Github.

# Learning Objectives

By the end of this milestone, you should:

-   Become familiar with your dataset of choosing
-   Select 4 questions that you would like to answer with your data
-   Generate a reproducible and clear report using R Markdown
-   Become familiar with manipulating and summarizing your data in tibbles using `dplyr`, with a research question in mind.

# Task 1: Choose your favorite dataset (10 points)

The `datateachr` package by Hayley Boyce and Jordan Bourak currently composed of 7 semi-tidy datasets for educational purposes. Here is a brief description of each dataset:

-   *apt_buildings*: Acquired courtesy of The City of Toronto's Open Data Portal. It currently has 3455 rows and 37 columns.

-   *building_permits*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 20680 rows and 14 columns.

-   *cancer_sample*: Acquired courtesy of UCI Machine Learning Repository. It currently has 569 rows and 32 columns.

-   *flow_sample*: Acquired courtesy of The Government of Canada's Historical Hydrometric Database. It currently has 218 rows and 7 columns.

-   *parking_meters*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 10032 rows and 22 columns.

-   *steam_games*: Acquired courtesy of Kaggle. It currently has 40833 rows and 21 columns.

-   *vancouver_trees*: Acquired courtesy of The City of Vancouver's Open Data Portal. It currently has 146611 rows and 20 columns.

**Things to keep in mind**

-   We hope that this project will serve as practice for carrying our your own *independent* data analysis. Remember to comment your code, be explicit about what you are doing, and write notes in this markdown document when you feel that context is required. As you advance in the project, prompts and hints to do this will be diminished - it'll be up to you!

-   Before choosing a dataset, you should always keep in mind **your goal**, or in other ways, *what you wish to achieve with this data*. This mini data-analysis project focuses on *data wrangling*, *tidying*, and *visualization*. In short, it's a way for you to get your feet wet with exploring data on your own.

And that is exactly the first thing that you will do!

1.1 Out of the 7 datasets available in the `datateachr` package, choose **4** that appeal to you based on their description. Write your choices below:

**Note**: We encourage you to use the ones in the `datateachr` package, but if you have a dataset that you'd really like to use, you can include it here. But, please check with a member of the teaching team to see whether the dataset is of appropriate complexity. Also, include a **brief** description of the dataset here to help the teaching team understand your data.

<!-------------------------- Start your work below ---------------------------->

1: cancer_sample
2: flow_sample
3: vancouver_trees
4: parking_meters

<!----------------------------------------------------------------------------->

1.2 One way to narrowing down your selection is to *explore* the datasets. Use your knowledge of dplyr to find out at least *3* attributes about each of these datasets (an attribute is something such as number of rows, variables, class type...). The goal here is to have an idea of *what the data looks like*.

*Hint:* This is one of those times when you should think about the cleanliness of your analysis. I added a single code chunk for you below, but do you want to use more than one? Would you like to write more comments outside of the code chunk?

<!-------------------------- Start your work below ---------------------------->

```{r}
### EXPLORE HERE ###
cancer_sample
flow_sample
vancouver_trees
parking_meters

glimpse(cancer_sample)
cancer_sample %>% group_by(diagnosis) %>% summarise(area_mean = mean(area_mean, na.rm = TRUE))

glimpse(flow_sample)
flow_sample %>% summarise(flow = range(flow, na.rm = TRUE))
flow_sample %>% summarise(month = mean(month, na.rm = TRUE))

glimpse(vancouver_trees)
summary(vancouver_trees)
vancouver_trees %>% summarise(diameter = mean(diameter,na.rm = TRUE))

glimpse(parking_meters)
parking_meters %>% summarise(latitude = range(latitude,na.rm = TRUE))
```

cancer_sample

1. This dataset contains 569 rows and 32 columns.

2. One column (diagnosis) is categorical and the other 31 are numerical.

3. The mean area_mean for benign tumors is 462.7902 and the mean area_mean for malignant tumors is 978.3764.

flow_sample

1. This dataset contains 218 rows and 7 columns.

2. The range of the flow variable is between 3.62 and 466.00.

3. The mean month is 5. 

vancouver_trees

1. This dataset has 146,611 rows and 20 columns.

2. This dataset has 12 categorical variables.

3. The mean diameter is 11.49016. 

parking_meters

1. This dataset has 10,032 rows and 22 columns.

2. This dataset has 20 categorical and 2 numerical columns.

3. The latitude range is from 49.21070 to 49.29329.

<!----------------------------------------------------------------------------->

1.3 Now that you've explored the 4 datasets that you were initially most interested in, let's narrow it down to 2. What lead you to choose these 2? Briefly explain your choices below, and feel free to include any code in your explanation.

<!-------------------------- Start your work below ---------------------------->
cancer_sample

* I understand the information in each column and I think I can formulate research questions using this dataset.

vancouver_trees

* I've worked with tree data before and also think I could formulate research questions using this dataset.

<!----------------------------------------------------------------------------->

1.4 Time for the final decision! Going back to the beginning, it's important to have an *end goal* in mind. For example, if I had chosen the `titanic` dataset for my project, I might've wanted to explore the relationship between survival and other variables. Try to think of 1 research question that you would want to answer with each dataset. Note them down below, and make your final choice based on what seems more interesting to you!

<!-------------------------- Start your work below ---------------------------->
**cancer_sample** 

* What is the relationship between symmetry_mean and cavity_mean in patients with benign tumors?

**vancouver_trees** 

* What is the relationship between neighborhood and diameter for Americana species?

<!----------------------------------------------------------------------------->

# Important note

Read Tasks 2 and 3 *fully* before starting to complete either of them. Probably also a good point to grab a coffee to get ready for the fun part!

This project is semi-guided, but meant to be *independent*. For this reason, you will complete tasks 2 and 3 below (under the **START HERE** mark) as if you were writing your own exploratory data analysis report, and this guidance never existed! Feel free to add a brief introduction section to your project, format the document with markdown syntax as you deem appropriate, and structure the analysis as you deem appropriate. Remember, marks will be awarded for completion of the 4 tasks, but 10 points of the whole project are allocated to a reproducible and clean analysis. If you feel lost, you can find a sample data analysis [here](https://www.kaggle.com/headsortails/tidy-titarnic) to have a better idea. However, bear in mind that it is **just an example** and you will not be required to have that level of complexity in your project.

# Task 2: Exploring your dataset (15 points)

If we rewind and go back to the learning objectives, you'll see that by the end of this deliverable, you should have formulated *4* research questions about your data that you may want to answer during your project. However, it may be handy to do some more exploration on your dataset of choice before creating these questions - by looking at the data, you may get more ideas. **Before you start this task, read all instructions carefully until you reach START HERE under Task 3**.

2.1 Complete *4 out of the following 8 exercises* to dive deeper into your data. All datasets are different and therefore, not all of these tasks may make sense for your data - which is why you should only answer *4*. Use *dplyr* and *ggplot*.

1.  Plot the distribution of a numeric variable.
2.  Create a new variable based on other variables in your data (only if it makes sense)
3.  Investigate how many missing values there are per variable. Can you find a way to plot this?
4.  Explore the relationship between 2 variables in a plot.
5.  Filter observations in your data according to your own criteria. Think of what you'd like to explore - again, if this was the `titanic` dataset, I may want to narrow my search down to passengers born in a particular year...
6.  Use a boxplot to look at the frequency of different observations within a single variable. You can do this for more than one variable if you wish!
7.  Make a new tibble with a subset of your data, with variables and observations that you are interested in exploring.
8.  Use a density plot to explore any of your variables (that are suitable for this type of plot).

2.2 For each of the 4 exercises that you complete, provide a *brief explanation* of why you chose that exercise in relation to your data (in other words, why does it make sense to do that?), and sufficient comments for a reader to understand your reasoning and code.

<!-------------------------- Start your work below ---------------------------->

### Histogram
The following code is used to display the distribution of the mean smoothness for all tumors.
```{r}
ggplot(cancer_sample, aes(smoothness_mean, ..density..)) +
  geom_histogram()
```

This is useful because it shows that the smoothness_mean variable follows a normal distribution with 0.100 being the most common value for this variable. 

### Boxplots
The following code is used to create a boxplot of the mean area of malignant and benign tumors.
```{r}
ggplot(cancer_sample, aes(diagnosis, area_mean, colour = diagnosis)) + geom_boxplot(width = 0.2)
```

The following code is used to create a boxplot of the mean concavity of malignant and benign tumors.
```{r}
ggplot(cancer_sample, aes(diagnosis, concavity_mean, colour = diagnosis)) + geom_boxplot(width = 0.2)
```

Boxplots are useful because they allow you to compare the values of a continuous variable between categorical variables. 
Boxplots allow for visualization of the minimum value, first quartile, median, third quartile, and maximum value. You can also see if any outliers are present. 
This dataset has one categorical variable and it is useful to visualize how a numeric variable differs based on tumor type.

### Geom_point
The following code is used to create a scatterplot to explore the relationship between the two continuous variables compactness_mean and concavity_mean for both benign and malignant tumors. 
```{r}
ggplot(cancer_sample, aes(x=compactness_mean, y=concavity_mean)) + 
  geom_point(aes(colour = diagnosis),
              size = 3,
              alpha = 0.8)
```

This is useful because this dataset contains several continuous variables so it is beneficial to observe the relationships between them and if the relationship is the same or different for malignant and benign tumors.

### Density Plot
The following code is used to create a density plot of texture_worst for both benign and malignant tumors.
```{r}
ggplot(cancer_sample, aes(x = texture_worst)) + 
  geom_density(aes(group=diagnosis, colour=diagnosis))
```

This density plot is useful because it displays that majority of the values for texture_worst are at approximately 27 for malignant tumors and 21 for benign tumors.
  
<!----------------------------------------------------------------------------->

# Task 3: Write your research questions (5 points)

So far, you have chosen a dataset and gotten familiar with it through exploring the data. Now it's time to figure out 4 research questions that you would like to answer with your data! Write the 4 questions and any additional comments at the end of this deliverable. These questions are not necessarily set in stone - TAs will review them and give you feedback; therefore, you may choose to pursue them as they are for the rest of the project, or make modifications!

<!--- *****START HERE***** --->

1. How many individuals with a benign tumor diagnosis have a symmetry_mean value less than 0.15?

2. What proportion of individuals with a malignant tumor diagnosis have an area_mean greater than 1200?

3. Are the values for area_worst greater among individuals with B or M diagnosis and by how much?

4. Is the relationship between concave_points_mean and concavity_mean the same between B and M diagnosis and what is the relationship?

# Task 4: Process and summarize your data (13 points)

From Task 2, you should have an idea of the basic structure of your dataset (e.g. number of rows and columns, class types, etc.). Here, we will start investigating your data more in-depth using various data manipulation functions.

### 1.1 (10 points)

Now, for each of your four research questions, choose one task from options 1-4 (summarizing), and one other task from 4-8 (graphing). You should have 2 tasks done for each research question (8 total). Make sure it makes sense to do them! (e.g. don't use a numerical variables for a task that needs a categorical variable.). Comment on why each task helps (or doesn't!) answer the corresponding research question.

Ensure that the output of each operation is printed!

**Summarizing:**

1.  Compute the *range*, *mean*, and *two other summary statistics* of **one numerical variable** across the groups of **one categorical variable** from your data.
2.  Compute the number of observations for at least one of your categorical variables. Do not use the function `table()`!
3.  Create a categorical variable with 3 or more groups from an existing numerical variable. You can use this new variable in the other tasks! *An example: age in years into "child, teen, adult, senior".*
4.  Based on two categorical variables, calculate two summary statistics of your choosing.

**Graphing:**

5.  Create a graph out of summarized variables that has at least two geom layers.
6.  Create a graph of your choosing, make one of the axes logarithmic, and format the axes labels so that they are "pretty" or easier to read.
7.  Make a graph where it makes sense to customize the alpha transparency.
8.  Create 3 histograms out of summarized variables, with each histogram having different sized bins. Pick the "best" one and explain why it is the best.

Make sure it's clear what research question you are doing each operation for!

<!------------------------- Start your work below ----------------------------->
### Question 1
1. How many individuals with a benign tumor diagnosis have a symmetry_mean value less than 0.15?

The following code is used to compute the number of observations for the categorical variable diagnosis.
```{r}
count(cancer_sample, diagnosis, wt = NULL, sort = FALSE, name = NULL)
```


Three histograms of the symmetry_mean variable for benign tumors are displayed with different sized bins.
```{r}
benign <- filter(cancer_sample, diagnosis == "B")
ggplot(benign, aes(symmetry_mean, ..density..)) +
  geom_histogram(bins = 30)
ggplot(benign, aes(symmetry_mean, ..density..)) +
  geom_histogram(bins = 50)
ggplot(benign, aes(symmetry_mean, ..density..)) +
  geom_histogram(bins = 10)
```

The second histogram with 50 sized bins is preferred because it best displays how the distribution changes at each symmetry_mean value. 

### Question 2
2. What proportion of individuals with a malignant tumor diagnosis have an area_mean greater than 1200?

The following table of only metrics for malignant tumors displays am additional column called "area_mean_category" which is a categorical variable made from the area_mean numeric variable. Small indicates an area less than 800, medium indicates an area between 800 and 1200, and large indicates an area greater than 1200.

```{r}
malignant <- filter(cancer_sample, diagnosis == "M")
malignant %>% summarise(area_mean = range(area_mean, na.rm = TRUE))
malignant %>% mutate(area_mean_category=cut(area_mean, breaks=c(361.6, 800, 1200, 2501.0), labels=c("small", "medium", "large")))
```

The following graph contains two geom layers and shows the distribution of the mean area for malignant tumors.

```{r}
ggplot(malignant, aes(x=area_mean)) +
  geom_histogram(aes(y=..density..),
                 colour = 1, fill = "blue") +
  geom_density()
```

### Question 3
3. Are the values for area_worst greater among individuals with a benign or malignant tumor diagnosis and by how much?

The following tables show the mean, range, median, and interquartile range of worst area for both benign and malignant tumors.

```{r}
cancer_sample %>% group_by(diagnosis) %>% summarise(area_worst = mean(area_worst, na.rm = TRUE))
cancer_sample %>% group_by(diagnosis) %>% summarise(area_worst = range(area_worst, na.rm = TRUE))
cancer_sample %>% group_by(diagnosis) %>% summarise(area_worst = median(area_worst, na.rm = TRUE))
cancer_sample %>% group_by(diagnosis) %>% summarise(area_worst = IQR(area_worst, na.rm = TRUE))
```

The following boxplot displays how worst area differs between benign and malignant tumors. The y-axis displaying area_worst is logarithmic. 

```{r}
ggplot(cancer_sample, aes(diagnosis, area_worst, colour = diagnosis)) + geom_boxplot(width = 0.2) + scale_y_log10()
```

### Question 4
4. Is the relationship between concave_points_mean and concavity_mean the same between B and M diagnosis and if so what is the relationship?

The following tables show the mean, range, median, and interquartile range of mean concave points for both benign and malignant tumors.

```{r}
cancer_sample %>% group_by(diagnosis) %>% summarise(concave_points_mean = mean(concave_points_mean, na.rm = TRUE))
cancer_sample %>% group_by(diagnosis) %>% summarise(concave_points_mean = range(concave_points_mean, na.rm = TRUE))
cancer_sample %>% group_by(diagnosis) %>% summarise(concave_points_mean = median(concave_points_mean, na.rm = TRUE))
cancer_sample %>% group_by(diagnosis) %>% summarise(concave_points_mean = IQR(concave_points_mean, na.rm = TRUE))
```

Below displays a scatterplot of the relationship between concave_points_mean on the x-axis and concavity_mean on the y-axis for both benign (pink) and malignant (blue) tumors. The alpha transparency has been changed to 0.2.

```{r}
ggplot(cancer_sample, aes(x=concave_points_mean, y=concavity_mean)) + 
  geom_point(aes(colour = diagnosis),
              size = 3,
              alpha = 0.2)
```
<!----------------------------------------------------------------------------->

### 1.2 (3 points)

Based on the operations that you've completed, how much closer are you to answering your research questions? Think about what aspects of your research questions remain unclear. Can your research questions be refined, now that you've investigated your data a bit more? Which research questions are yielding interesting results?

<!-------------------------- Start your work below ---------------------------->
**Question 1**

1. How many individuals with a benign tumor diagnosis have a symmetry_mean value less than 0.15?

* Computting the number of observations for the diagnosis variable allowed me to determine that there are 357 individals with benign tumors but this does not tell me anything about the symmetry_mean variable.

* Creating three histograms with different sized bins allowed me to visualize the symmetry_mean distribution for benign tumors. This allows me to estimate the number of benign tumors with a symmetry_mean value of less than 0.15.

**Question 2**

2. What proportion of individuals with a malignant tumor diagnosis have an area_mean greater than 1200?

* Filtering the dataset to malignant tumors and categorizing the numeric area_mean variable was very useful because I can count the number of "Large" tumors in the area_mean_category column since these are tumors with a mean area greater than 1200.

* The graph with two geom layers allowed me to visualize how many malignant tumors have an area greater than 1200 but since the y-axis displays density I cannot estimate the number of malignant tumors with an area greater than 1200.

**Question 3**

3. Are the values for area_worst greater among individuals with a benign or malignant tumor diagnosis and by how much?

* Computing the mean, range, median, and interquartile range of area_worst for both malignant and benign tumors was useful because the range allowed me to determine malignant tumors have greater maximum area_worst value compared to benign tumors. 

* Creating a boxplot of worst area of malignant and benign tumors allowed me to visualize that malignant tumors generally have a greater worst area compared to benign tumors but there is some overlap. I still don't know how much greater the area_worst values are for malignant tumors. 

**Question 4**

4. Is the relationship between concave_points_mean and concavity_mean the same between benign and malignant tumors and what is the relationship?

* Computing the mean, range, median, and interquartile range of concave_points_mean for both malignant and benign tumors not very useful for understanding the relationship between concave_points_mean and concavity_mean.

* Creating a scatterplot of the relationship between concave_points_mean and concavity_mean for malignant and benign tumors was useful because there appears to be a linear relationship for both malignant and benign tumors. There are more outliers for benign tumors however. 

<!----------------------------------------------------------------------------->

### Attribution

Thanks to Icíar Fernández Boyano for mostly putting this together, and Vincenzo Coia for launching.
